---
title: "Random Intercept Cross Lagged Panel Models in {lavaan}"
description: |
  A tutorial on fitting RICLPMs.
author: Christopher M. Loan
date: 2021-09-15
output:
  distill::distill_article:
    self_contained: false
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lavaan)
```

While getting my Masters Degree in Prevention Science, I was working with some psychosocial variables that were both known to contribute to substance use among adolescents. There was mixed evidence in the literature about which of these had the greater and earliest effect. This is a tough thing to disentangle, and (almost) exclusively answerable with Structural Equation Modeling (SEM), due to the time-lagged, multiple-outcome nature of the question. 

After digging through the methodological possibilities, the research team settled on the Random Intercept Cross Lagged Panel Model (RICLPM; [Hamaker, Kuiper, & Grasman (2015)](https://www.researchgate.net/profile/Ellen-Hamaker/publication/274262847_A_Critique_of_the_Cross-Lagged_Panel_Model/links/5b80154c92851c1e122f351f/A-Critique-of-the-Cross-Lagged-Panel-Model.pdf)). A recent publication expands these methods [(Mulder & Hamaker, 2020)](https://www.tandfonline.com/doi/pdf/10.1080/10705511.2020.1784738). If you're interested in more technical details than simply fitting these models, I'd start with these. Here's [some good code](https://jeroendmulder.github.io/RI-CLPM/) to supplement these models, put out by the authors themselves. 

I'm going to first simulate some data. I'll be using the same data that I used in my [growth and parallel process model post](https://www.christopherloan.com/blog/growth-ppm/). 

The goal of this post is to fit an RICLPM with an outcome predicted by the cross-lagged panel variables (a.k.a. the within-person variables). I will then show how to use bootstrapping to calculate a mediated effect through the panel portion to the outcome

```{r}

growth_mod <- 
  '

## intercept & slope growth terms for X
iX =~ 1*x1 + 1*x2 + 1*x3 + 1*x4 + 1*x5
sX =~ 0*x1 + 1*x2 + 2*x3 + 3*x4 + 4*x5

## intercept, slope, & quadratic terms for Y
iY =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5 
sY =~ 0*y1 + 1*y2 + 2*y3 + 3*y4 + 4*y5 
#qY =~ 0*y1 + 1*y2 + 4*y3 + 9*y4 + 16*y5

# ## set variances
sY ~~ 2*sY
sX ~~ 3*sX

## set intercepts
iX ~ 3*1
sX ~ 1.6*1
sY ~ -1*1
#qY ~ -1.5*1

sY ~ 2*predictor

outcome ~ 2*iX + 3*sY 

  '
```

```{r }
sim_growth_dat <- 
  simulateData(model = growth_mod, 
               model.type = "growth", 
               seed = 82020, 
               orthogonal = F,
               auto.cov.y = T, 
               auto.var = T, 
               sample.nobs = 2000
               )

```

```{r}
mod1 <- '
  # Create between components (random intercepts)
  RIx =~ 1*x1 + 1*x2 + 1*x3 + 1*x4 + 1*x5
  RIy =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
  
  # Estimate the variance and covariance of the random intercepts. 
  RIx ~~ RIx
  RIy ~~ RIy
  RIx ~~ RIy
  
  # Regression of time-invariant outcome z2 on within components.
  outcome ~ wx1 + wx2 + wx3 + wx4 + wx5 + wy1 + wy2 + wy3 + wy4 + wy5
  outcome ~~ outcome # Residual variance z2
  
  # Create within-person centered variables
  wx1 =~ 1*x1
  wx2 =~ 1*x2
  wx3 =~ 1*x3 
  wx4 =~ 1*x4
  wx5 =~ 1*x5
  wy1 =~ 1*y1
  wy2 =~ 1*y2
  wy3 =~ 1*y3
  wy4 =~ 1*y4
  wy5 =~ 1*y5
  
  # # Regression of observed variables on z1 (constrained). 
  #x1 + x2 + x3 + x4 + x5 ~ s1*z1 # Constrained over time.
  y1 + y2 + y3 + y4 + y5 ~ s2*predictor # Constrained over time.
  
  # Estimate the lagged effects between the within-person centered variables.
  wx2 + wy2 ~ wx1 + wy1
  wx3 + wy3 ~ wx2 + wy2
  wx4 + wy4 ~ wx3 + wy3
  wx5 + wy5 ~ wx4 + wy4
  
  # Estimate the covariance between the within-person centered variables at the first wave. 
  wx1 ~~ wy1 # Covariance
  
  # Estimate the covariances between the residuals of the within-person centered variables (the innovations).
  wx2 ~~ wy2
  wx3 ~~ wy3
  wx4 ~~ wy4
  wx5 ~~ wy5
  
  # Estimate the (residual) variance of the within-person centered variables.
  wx1 ~~ wx1 # Variances
  wy1 ~~ wy1 
  wx2 ~~ wx2 # Residual variances
  wy2 ~~ wy2 
  wx3 ~~ wx3 
  wy3 ~~ wy3 
  wx4 ~~ wx4 
  wy4 ~~ wy4 
  wx5 ~~ wx5
  wy5 ~~ wy5
'

fit1 <- lavaan(
  mod1, 
  data = sim_growth_dat, 
  meanstructure = T, 
  int.ov.free = T
  ) 

summary(fit1, standardized = T)
```

THIS IS ALL DIRECTLY COPIED AND SHOULD BE REWORDED

```{r}
mod2 <- '
  # Create between components (random intercepts)
  RIx =~ 1*x1 + 1*x2 + 1*x3 + 1*x4 + 1*x5
  RIy =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
  
  # Estimate the variance and covariance of the random intercepts. 
  RIx ~~ RIx
  RIy ~~ RIy
  RIx ~~ RIy
  
  # Regression of time-invariant outcome z2 on within components.
  outcome ~ wx1 + wx2 + wx3 + wx4 + wx5 + wy1 + wy2 + wy3 + wy4 + wy5
  outcome ~~ outcome # Residual variance z2
  
  # Create within-person centered variables
  wx1 =~ 1*x1
  wx2 =~ 1*x2
  wx3 =~ 1*x3 
  wx4 =~ 1*x4
  wx5 =~ 1*x5
  wy1 =~ 1*y1
  wy2 =~ 1*y2
  wy3 =~ 1*y3
  wy4 =~ 1*y4
  wy5 =~ 1*y5
  
  # Regression of observed variables on z1 (constrained). 
  y1 + y2 + y3 + y4 + y5 ~ s2*predictor # Constrained over time.
  
  # Estimate the lagged effects between the within-person centered variables.
  
  wx2 + wy2 ~ wx1 + wy1
  wx3 + wy3 ~ wx2 + wy2
  wx4 + wy4 ~ wx3 + wy3
  wx5 + wy5 ~ wx4 + wy4
  
  # Estimate the covariance between the within-person centered variables at the first wave. 
  wx1 ~~ wy1 # Covariance
  
  # Estimate the covariances between the residuals of the within-person centered variables (the innovations).
  wx2 ~~ wy2
  wx3 ~~ wy3
  wx4 ~~ wy4
  wx5 ~~ wy5
  
  # Estimate the (residual) variance of the within-person centered variables.
  wx1 ~~ wx1 # Variances
  wy1 ~~ wy1 
  wx2 ~~ wx2 # Residual variances
  wy2 ~~ wy2 
  wx3 ~~ wx3 
  wy3 ~~ wy3 
  wx4 ~~ wx4 
  wy4 ~~ wy4 
  wx5 ~~ wx5
  wy5 ~~ wy5
'

fit2 <- lavaan(
  mod2, 
  data = sim_growth_dat, 
  meanstructure = T, 
  int.ov.free = T
  ) 

summary(fit2, standardized = T)
```
