---
title: "APA 2021 Supplemental Material"
description: 
  "Supplmental Material for the Application of Generalized Linear Mixed-Effects Model Regression Trees with Multi-Modal Data poster presented at APA 2021, co-authored with Matthew C. Graham, Edwin M. Zorilla, Idalis Villanueva Alarcon, Jenefer Husman, & Keith Zvoch"
author:
  - name: Christopher M. Loan
    url: {}
date: 2021-08-03
categories:
  - Collaboration
  - Conference
  - Machine Learning
  - Multi-Level Modeling
  - MLM
  - Plotting
  - Simulated Data
  - Tutorial 
output:
  distill::distill_article:
    self_contained: false
    code_folding: show code
    toc: true
draft: true
---

The purpose of these supplemental materials is to walk users through using the `{glmertree}` package and provide references/supplemental reading for our poster. Refer to the Table of Contents (or scroll down) if you're just here for references. 

For readability, code is included but can be hidden by clicking `show code`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

I default to writing code in the style of the [`{tidyverse}`](https://www.tidyverse.org/), so please see that if you're unfamiliar with this style of coding. Essentially, you can "pipe" objects—typically data—from one function to another. If you're using `R` these days, you've probably seen it, even if you don't use it. 

So everything will be done with `{tidyverse}` functions and the`{glmertree}` package. I'll compare some `{glmertree}` results to `{lme4}`, but this is loaded via `{glmertree}`  anyways. I'll try and use `::` operator to "namespace" functions that aren't from the aforementioned packages, so you know what to install and from where it's originating

# libraries

```{r libs}
library(glmertree)
library(tidyverse)
```

# simulating data: skip this if you just want `{glmertree}` model specification

In a perfect world, I'd be able to share the data with you and use the same data used on the poster without anything bad happening; however, we have data privacy for a reason. I still want to show you how to specify these models, and I want you to have the option to follow along. The supplemental data can be simulated on your machine if you follow the instructions on [another blog post of mine](www.christopherloan.com/blog/simulating-2-level-data-for-apa-2021-supplemental/). 

If you don't care about the simulation and just want to see model specification, you're in the right place.   

The short of it: I simulate some with similar size level 1 and level 2 units as our poster, but we're going to omit the cross-classified design just to make this process a bit simpler to follow.

```{r functions, include = FALSE}
simulate_interaction_w_2_levels <-
  function(
    n, ## number level 1 units
    j, ## number level 2 units
    intercept_lv1, ## intercept at level 1
    main_x, ## main effect of x
    main_z, ## main effect of z
    interaction, ## interaction of x and z
    residual_var_sd_lv1, ## standard deviation of residuals at level 1
    random_int_mean_lv2, ## mean of random intercept at level 2
    random_int_sd_lv2, ## standard deviation of random intercept at level 2,
    start_seed = 123
  ){
    
    ## placeholder for level 2 outcomes
    outcomes_j <- vector("list", j)
    
    ## for each variable, make list 
    ## as long as level 2
    ## fill each element with a list of level 1 outcomes
    
    
    x <- vector("list", j) 
    z <- vector("list", j) 
    w <- vector("list", j) 
    
    ## School distribution (intercept at level 2)
    ## Standard deviation of random intercept at level 2
    set.seed(start_seed)
    
    a_j <- rnorm(j, random_int_mean_lv2, random_int_sd_lv2)
    
    for(i in 1:j) {
      
      ## make a level 1 predictor variable:
      ## set multiple seeds that change each iteration
      ## prevents identical cases with seed
      
      set.seed(start_seed+i)
      x[[i]] <- rnorm(n)
      set.seed(start_seed-i)
      z[[i]] <- rnorm(n)
      set.seed(-start_seed + i)
      outcomes_j[[i]] <- 
        rnorm(
          n, 
          intercept_lv1 + 
            interaction*z[[i]]*x[[i]] + 
            main_x*x[[i]] + 
            main_z*z[[i]] + 
            a_j[i], 
          ## standard deviation of residual variance
          residual_var_sd_lv1
        )
      set.seed(start_seed*197+197*i)
      w[[i]] <- rnorm(n)
    }
    
    outcomes_df <- 
      data.frame(
        id = rep(1:j, each = n),
        x = unlist(x),
        z = unlist(z),
        outcome = unlist(outcomes_j),
        w = unlist(w)
      )
    
    return(outcomes_df)
  }
```

```{r plot_theme, include=F}
project_theme <- 
  theme(
    axis.line = element_line(size = 3, lineend = 'round'), 
    legend.position = 'none',
    plot.title.position = 'plot', 
    panel.grid.minor = element_blank(),
    axis.title.x = element_text(size = 18),
    axis.text = element_text(size = 16),
    axis.title.y = element_text(size = 20),
    plot.title = element_text(size = 25),
    strip.text = element_text(size = 25),
  )
```

We have 15 level 1 units (number of question in our poster), 71 level 2 units (students in our poster). The level 1 intercept of the outcome is 4.25 with a main effect of 1.25 for x, 2.00 for z, and 0.75 for their interaction. The function also makes a nuisance variable `w`, which does not contribute to the outcome at all. 

```{r simulate}
sim_dat <- 
  simulate_interaction_w_2_levels(
    n = 15, ## number level 1 units at each level 2
    j = 71, ## number level 2 units
    intercept_lv1 = 4.25, ## intercept at level 1
    main_x = 1.25, ## main effect of x
    main_z = 2.00, ## main effect of z
    interaction = 0.75, ## interaction of x and z
    residual_var_sd_lv1 = 2.00, ## standard deviation of residuals at level 1
    random_int_mean_lv2 = 0, ## mean of random intercept at level 2
    random_int_sd_lv2 = 1.00, ## standard deviation of random intercept at level 2,
    start_seed = 123 ## ensure you can reproduce 
  ) %>% 
  tibble()
```

## MLM parameters with this data, via `{lme4}`

```{r lmer}
confirmation_lmer <- 
  lmer(
    data = sim_dat, 
    formula = outcome ~ x * z + w + # equivalent to x + z + x : z
      (1 | id)
  )
summary(confirmation_lmer)
```

We see how closely the MLM estimates identify the simulated effects. We see small standard errors and large t-values for these estimates, too. As we'd like to see, `w` does not have significant associations with the outcome.

# glmertrees

Specifying models with `{glmertree}` is very similar to `{lme4}`. 

The formula is slightly different; rather than adding random effects, you specify the model as follows: `forumla = fixed effects | random effects | splitting variables`

Another consideration is that you're using parameter instability tests in addition to MLMs, so you want to specify the `id` variable to the `cluster = id` argument as well. 

Let's run a very exploratory model here. This is everything but the kitchen sink: an intercept-only model, with random intercept for each level 2 unit, and allowing splits to be made by any of the variables in the data.

```{r tree 1}
tree_1 <- 
  lmertree(
    data = sim_dat, 
    formula = 
      outcome ~ 1 | (1 | id) | x + z + w, 
    cluster = id, 
  )
```

Let's look at plotted results first; adding `ask = FALSE` to your code lets you see the plotted decision tree and the random effects caterpillar plot without having to click `<return>` on your keyboard.

```{r plot tree 1, fig.width = 8, fig.height = 6}
plot(tree_1, ask = FALSE)
```

Our model correctly identified the influence of `z` and `x`, but we should probably   

```{r tree 2}
tree_2 <- 
  lmertree(
    data = sim_dat, 
    formula = 
      outcome ~ 1 | (1 | id) | x + z + w, 
    cluster = id, 
    minsize = 0.20*nrow(sim_dat),
    bonferroni = T,
    alpha = 0.001
  )
```

```{r plot tree 2, fig.width = 8, fig.height = 6}
plot(tree_2$tree, ask = F)
```

```{r tree 3}
tree_3 <- 
  lmertree(
    data = sim_dat, 
    formula = 
      outcome ~ x * z | (1 | id) | w, 
    cluster = id, 
    #minsize = 0.50*nrow(sim_dat),
    alpha = 0.001
  )
```

```{r plot tree 3, fig.width = 8, fig.height = 6}
plot(tree_3$tree, ask = F)
```

```{r}
summary(tree_3$lmer)
```

```{r}
summary(tree_3$tree)
```

